#### Run the general summary functions
```{r, echo = T, warning = F, message = F}

# The Task:
  # For each CLASS of chemicals (e.g. Aroclors by method SW8082A)
  # a summary sample result is generated that complies with 
  # agency specific summation rule, which usually are concerned with 
  # how to account for results below the analytical limit of detection (DL).
  
  # A dblResult can be a number OR NA
  # every dblResult has an associated string flag OR NA
  # If a result is NA, we consider it as a concentration of 0, 1/2DL, or DL
  
  # We also keep track of all the unique flags that went into a summary
  
  # For book keeping we must keep track of:
    # (1) the number of dblresults of each CLASS, 
    # (2) the number of NA dblresult below the Detection Limit.
    # Additionally, if the number of results per sample, within any sample CLASS differ, 
    # that is an unexpected result, indicating an incomplete or unbalanced input file
    # It should cause the routine to stop. an easy test would be:
    # group_by(txtSample, CLASS) %>% 
      # summarise(n=n()) %>% 
      # ungroup() %>% 
      # group_by(CLASS) %>% 
      # summarise(Failure = (var(n)>0))


# BELOW ARE FUNCTIONS for reducing the data qualification flags 
  # to contain only unique information. 
  # Lab data comes with alphanumeric qualifications.
  # When summing multiple analytes, we must generate a summary flag 
  # consisting of the all the unique flags in the group plus a big "T"
  # to designate the result an aggregate Total.
  # By defining a set of concise helper functions 
  # this task can slide alongside the task 
  # of summarising analytical result values if
  # placed directly into a dplyr group_by() %>% summarise() 
  # data manipulation.
  # Note the insertion of this task in 
  # the final line of the summarise() function 
  # in the general_analytical_summary() function


# given any string turn return an alphabetal list of unique single charcters
get_unique_chars         <- function(x) {sort(unique(strsplit(x, "")[[1]]))}

# collapse a vector of strings toa single string
collapse_list_of_chars   <- function(x) {paste(x, collapse = "")}

# append a special character flag to a strin
append_special_flag      <- function(x, special = "T") {paste(x,special, sep = "")}

# remove all instances of a special character from a string
remove_special_flag      <- function(x, special = "NA") {
  stringr::str_replace_all(x, pattern =special, replacement = "")
}

# return all all unique flags
return_all_unique_flags <- function(x) {
  result <- collapse_list_of_chars(sort(unique(x))) %>%
    remove_special_flag("NA") %>%
    remove_special_flag("U") %>% 
    get_unique_chars() %>%
    collapse_list_of_chars() %>%
    append_special_flag("T")
  return(result) 
}

#' general_analytical_summary
#'
#' General purpose summary function for analytical results.
#' 
#' @param df input data.frame object must contain columns:
#' CLASS, txtConstituent, txtSampleID, dblResult, dblLimit, txtQual
#'
#' @return df.sums output data.frame
general_analytical_summary <- function(df){
  df.sums <- df %>% select(CLASS,
                           txtConstituent,
                           txtSampleID, 
                           dblResult, 
                           dblLimit, 
                           txtQual) %>% 
    mutate(dblResultzeroND = ifelse(is.na(dblResult), 0*dblLimit,   dblResult)) %>%
    mutate(dblResulthalfND = ifelse(is.na(dblResult), 0.5*dblLimit, dblResult)) %>%
    mutate(dblResultfullND = ifelse(is.na(dblResult), 1*dblLimit,   dblResult)) %>% 
    group_by(CLASS, txtSampleID) %>%
    summarise(sumResult_0ND    = sum(dblResultzeroND, na.rm=TRUE), 
              sumResult_halfND = sum(dblResulthalfND, na.rm=TRUE),
              sumResult_fullND = sum(dblResultfullND, na.rm=TRUE),
              maxLimit         = max(dblLimit), 
              n                = n(),
              nd               = sum(is.na(dblResult)),
              txtQual       = return_all_unique_flags(txtQual) ) %>%
    mutate(`ND_zero` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_0ND), 
           `ND_half` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_halfND),
           `ND_full` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_fullND))
  return(df.sums)
}


#' general_analytical_summary_PDI
#'
#' General purpose summary function for analytical results.
#' 
#' @param df input data.frame object must contain columns:
#' CLASS, txtConstituent, txtSampleID, dblResult, dblLimit, txtQual
#'
#' @return df.sums output data.frame
# this is a version of the general analytical summary meant to comply strictly with 
    # page 3. | Appendix E: 2018 PDI Surface Sediment Database Description
    # "Calculated totals are the sum of all detected results and one half of the highest reporting
    # detection limit for the results reported as not detected (ND) in a sample."
general_analytical_summary_PDI <- function(df){
  df.maxLimits <- df %>% select(CLASS,
                              txtConstituent,
                              txtSampleID,
                              dblLimit,
                              dblResult) %>%
    mutate(DL = is.na(dblResult)) %>%                      # This labels ND samples as DL = TRUE
    group_by(CLASS, txtSampleID, DL) %>%                   # This groups all the ND samples together per CLASS,txtSampleID
    summarise(dblmaxLimitSample = max(dblLimit, na.rm= T), # maxLimit calculates highest limits among NDs, 
              totalNDs = n()) %>%     
    filter(DL == TRUE) %>% ungroup()  # this filters so that we only have maxLimit from the samples with atleaast one ND
  
  df.sums <- df %>% select(CLASS,
                           txtConstituent,
                           txtSampleID, 
                           dblResult, 
                           dblLimit, 
                           txtQual) %>% 
    left_join(df.maxLimits) %>%
    # page 3. | Appendix E: 2018 PDI Surface Sediment Database Description
    # "Calculated totals are the sum of all detected results and one half of the highest reporting
    # detection limit for the results reported as not detected (ND) in a sample."
    mutate(dblResultzeroND = ifelse(is.na(dblResult), 0*dblmaxLimitSample,   dblResult)) %>%
    mutate(dblResulthalfND = ifelse(is.na(dblResult), 0.5*dblmaxLimitSample, dblResult)) %>%
    mutate(dblResultfullND = ifelse(is.na(dblResult), 1*dblmaxLimitSample,   dblResult)) %>% 
    group_by(CLASS, txtSampleID) %>%
    summarise(sumResult_0ND    = sum(dblResultzeroND, na.rm=TRUE), 
              sumResult_halfND = sum(dblResulthalfND, na.rm=TRUE),
              sumResult_fullND = sum(dblResultfullND, na.rm=TRUE),
              maxLimit         = max(dblLimit), 
              n                = n(),
              nd               = sum(is.na(dblResult)),
              txtQual       = return_all_unique_flags(txtQual) ) %>%
    mutate(`ND_zero` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_0ND), 
           `ND_half` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_halfND),
           `ND_full` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_fullND))
  return(df.sums)
  
}



#' output_EDD_from_sums_and_ref
#'
#' @param x.sum a data.frame with column names
#'   "txtSampleID", 
#'   "sumResult_0ND", 
#'   "sumResult_halfND", 
#'   "sumResult_fullND", 
#'   "maxLimit", 
#'   "n", 
#'   "nd", 
#'   "ND_zero", 
#'   "ND_half", 
#'   "ND_full", 
#'   "txtQual", 
#'   "CLASS"
#' @param x.ref a data.frame with other sample information which will be joined. It must contain the following headers to avoid an error
#'    txtProjectName
#'    txtEvent
#'    txtLabID
#'    dtmSampleDate
#'    txtAnalysisMethod
#'    txtConstiuent
#'    txtPCode
#'    txtSampleID
#'    dtmExtractedDate
#'    dtmAnalyzedDate
#'    intDilution
#'    dblResult
#'    txtUnits
#'    txtQual
#'    maxLimit
#'    txtQA
#'    txtComments
#'
#' @return x3 - data.frame with summation results in EDD format
output_EDD_from_sums_and_ref<- function(x.sum, x.ref, my_signif = 2){
  
  ref_info <- x.ref %>% group_by(txtSampleID, CLASS) %>% slice(1) %>% 
    select(-txtConstituent, -txtCASID, -txtQual, -dblLimit, -dblResult)
  
  x1 <- x.sum %>% select(CLASS, txtSampleID, ND_zero, ND_half, ND_full,txtQual, maxLimit) %>%
    gather(summation, dblResult, -CLASS, -txtSampleID, -txtQual, -maxLimit) %>% 
    mutate(txtConstiuent = paste0(CLASS, "_", summation)) %>% ungroup()
  
  x2 <- x1 %>% left_join(ref_info)
  
  x3 <- x2 %>% transmute(PRJ_NAME  = txtProjectName,
                         EVENT     = txtEvent,
                         LAB_ID    = txtLabID,
                         SAMPLE_DT = dtmSampleDate,
                         ANALYSIS_METH  = txtAnalysisMethod,
                         CONSTITUENT    = txtConstiuent,
                         CAS_ID         = NA,
                         P_CODE         = txtPCode,
                         SAMPLE_ID      = txtSampleID,
                         EXTRACTED_DATE = dtmExtractedDate,
                         ANALYZED_DT    = dtmAnalyzedDate,
                         DILUTION       = intDilution,
                         RESULT         = dblResult,
                         UNITS          = txtUnits,
                         QUAL           = txtQual,
                         LIMIT          = maxLimit,
                         QA             = txtQA,
                         COMMENTS       = txtComments)
  # apply significant figures to result
  x3$RESULT <- signif(x3$RESULT, my_signif)
  #change posix date format to MM/DD/YYY hh:mm character
  x3$SAMPLE_DT<-as.character(x3$SAMPLE_DT, format='%m/%d/%y %H:%M')
  x3$EXTRACTED_DATE<-as.character(x3$EXTRACTED_DATE, format='%m/%d/%y %H:%M')
  x3$ANALYZED_DT<-as.character(x3$ANALYZED_DT, format='%m/%d/%y %H:%M')
  #change NAs and NA characters to blanks
  x3[is.na(x3)] <- ""
  return(x3)
}
```

#### Setup a Tidy Environment
```{r, echo = T, warning = F, message = F}
require(dplyr)
require(knitr)
require(tidyr)
require(RODBC)
```

# PCB Aroclors according to PDI Summation Rules

#### Load Data
```{r, echo = T, warning = F, message = F}
db   <- "L:/DTNA/PGG_SedDB_2018_SQL.mdb"
query_string = "SELECT * FROM ExR_2bSUM_All"
output_suffix = "CALC_PDI_AROCLOR_2018_10_16.txt"  
con2 <- odbcConnectAccess(db)  ###Make sure dbase is closed####
data <- sqlQuery(con2, query_string) 
data <- data %>% mutate_if(is.factor, as.character)
close(con2)
```

#### Subset the data and avoid any result with "T" flag
```{r, echo = T, warning = F, message = F}
A <- data %>% filter(txtAnalysisMethod == "SW8082A", !grepl(txtQual, pattern = "T")) 
#### Subset the data
```

#### Assign a CLASS or CLASSes to summarize over
```{r, echo = T, warning = F, message = F}
# Assign a class, for Aroclors you might choose "Total Aroclors" or "TotalSW8082A"
# If you want to do more complex summations, you can do by assigning as many CLASSes as
# you like.
A$CLASS <-"CALC_PDI_Aroclors"
```

### Run the general analytical summary function (PDI VERSION)
```{r, echo = T, warning = F, message = F}
S = general_analytical_summary_PDI(A)
R <- output_EDD_from_sums_and_ref(x.sum = S, x.ref = A) 
```

### Preview Output
```{r, echo = T, warning = F, message = F}
head(R)
```

### Write output to txt file
```{r, echo = T, warning = F, message = T}
db_name <- stringr::str_replace(db, "L:/", "") %>%
  stringr::str_replace("/", "_") %>% 
  stringr::str_replace(".mdb", "_")
message( sprintf("Writing to file: %s",paste0(db_name, output_suffix)))
write.table(R, 
            file = paste0(db_name, output_suffix), 
            quote = F, 
            row.names = F, 
            sep= "|")
```















# PCB Aroclors


This workflow is approrpiate for analtical summaries that don't require a toxicity adjustment factor.
The heavy lifting should be in the resuable functions above, 
This greatly simplies the entire work flow and encourages code resusability.

#### Load Data
```{r, echo = T, warning = F, message = F}
db   <- "L:/DTNA/PGG_SedDB_2018_SQL.mdb"
query_string = "SELECT * FROM ExR_2bSUM_All"
output_suffix = "SUMAROCLOR_2018_10_13.txt"  
con2 <- odbcConnectAccess(db)  ###Make sure dbase is closed####
data <- sqlQuery(con2, query_string) 
data <- data %>% mutate_if(is.factor, as.character)
close(con2)
```

#### Subset the data and avoid any result with "T" flag
```{r, echo = T, warning = F, message = F}
A <- data %>% filter(txtAnalysisMethod == "SW8082A", !grepl(txtQual, pattern = "T")) 
#### Subset the data
```

#### Assign a CLASS or CLASSes to summarize over
```{r, echo = T, warning = F, message = F}
# Assign a class, for Aroclors you might choose "Total Aroclors" or "TotalSW8082A"
# If you want to do more complex summations, you can do by assigning as many CLASSes as
# you like.
A$CLASS <-"Total Aroclors"
```

### Run the general analytical summary function
```{r, echo = T, warning = F, message = F}
S = general_analytical_summary(A)
R <- output_EDD_from_sums_and_ref(x.sum = S, x.ref = A) 
```

### Preview Output
```{r, echo = T, warning = F, message = F}
head(R)
```

### Write output to txt file
```{r, echo = T, warning = F, message = T}
db_name <- stringr::str_replace(db, "L:/", "") %>%
  stringr::str_replace("/", "_") %>% 
  stringr::str_replace(".mdb", "_")
message( sprintf("Writing to file: %s",paste0(db_name, output_suffix)))
write.table(R, 
            file = paste0(db_name, output_suffix), 
            quote = F, 
            row.names = F, 
            sep= "|")
```


# DDD, DDT, and DDE

#### Load Data
```{r, echo = T, warning = F, message = F}
db   <- "L:/DTNA/PGG_SedDB_2018_SQL.mdb"
query_string = "SELECT * FROM ExR_2bSUM_All"
output_suffix = "SUMDDdte_2018_10_13.txt"  
con2 <- odbcConnectAccess(db)  ###Make sure dbase is closed####
data <- sqlQuery(con2, query_string) 
data <- data %>% mutate_if(is.factor, as.character)
close(con2)
```

#### Subset the data and avoid any result with "T" flag
```{r, echo = T, warning = F, message = F}

DD <- data %>% mutate_if(is.factor, as.character) %>% 
  filter(txtConstituent %in% c("o,p'-DDD", "4,4'-DDD", "o,p-DDE","4,4'-DDE", "o,p'-DDT",  "4,4'-DDT"),
         !grepl(txtQual, pattern = "T"))
         
         

```

#### Assign a CLASS or CLASSes to summarize over
```{r, echo = T, warning = F, message = F}
# Assign a class, for Aroclors you might choose "Total Aroclors" or "TotalSW8082A"
# If you want to do more complex summations, you can do by assigning as many CLASSes as
# you like.

DDdte <- DD %>% mutate(CLASS = 
                      ifelse(txtConstituent %in% c("o,p'-DDD", "4,4'-DDD"), "Total DDD", 
                        ifelse(txtConstituent %in% c("o,p'-DDE", "4,4'-DDE"), "Total DDE", 
                          ifelse(txtConstituent %in% c("o,p'-DDT", "4,4'-DDT"), "Total DDT", NA)
                              )
                            )
                    )




```

#### Run the general analytical summary function
```{r, echo = T, warning = F, message = F}
S = general_analytical_summary(DDdte)
R <- output_EDD_from_sums_and_ref(x.sum = S, x.ref = DDdte) 
```

#### Preview Output
```{r, echo = T, warning = F, message = F}
head(R)
```

#### Write output to .txt file
```{r, echo = T, warning = F, message = T}
db_name <- stringr::str_replace(db, "L:/", "") %>%
  stringr::str_replace("/", "_") %>% 
  stringr::str_replace(".mdb", "_")
message( sprintf("Writing to file: %s",paste0(db_name, output_suffix)))
write.table(R, 
            file = paste0(db_name, output_suffix), 
            quote = F, 
            row.names = F, 
            sep= "|")
```


# Total DDx

#### To show how simplified the work flow, here is Total DDx in ONE CHUNK
```{r, echo = T, warning = F, message = T}
output_suffix = "SUMDDx_2018_10_13.txt"
DDx <- DD %>% mutate(CLASS = "Total DDx")
S = general_analytical_summary(DDx)
R <- output_EDD_from_sums_and_ref(x.sum = S, x.ref = DDx)
db_name <- stringr::str_replace(db, "L:/", "") %>%
  stringr::str_replace("/", "_") %>% 
  stringr::str_replace(".mdb", "_")
message( sprintf("Writing to file: %s",paste0(db_name, output_suffix)))
write.table(R, 
            file = paste0(db_name, output_suffix), 
            quote = F, 
            row.names = F, 
            sep= "|")

```


## TOTAL PAH
```{r, echo = T, warning = F, message = T}
# Total PAHs—Total polycyclic aromatic hydrocarbons (PAHs) are the sum of the individual LPAHs and HPAHs.

# FROM PAGE A3-12 of Portland Harbor RI SCRA DATABASE AND DATA MANAGEMENT

#Total LPAHs—Sum of acenaphthene, anthracene, biphenyl, 2,6-dimethylnaphthalene, fluorene, 1-methylnaphthene, 2-methylnapthene, 1-methylphenanthrene, phenanthrene, naphthalene, where two or more are measured.

#Total HPAHs—Sum of benzo(a)anthracene, benzo(a)pyrene, benzo(e)pyrene, chrysene, dibenzo(a,h)anthracene, fluoranthene, perylene, and pyrene, where two or more are measured.

pahs <- c("2-Methylnaphthalene", #LPAH1
          "Acenaphthene",        #LPAH2
          "Acenaphthylene",      #LPAH3
          "Anthracene",          #LPAH4
          "Fluorene",            #LPAH5
          "Naphthalene",         #LPAH6
          "Phenanthrene",        #LPAH7
          "Fluoranthene",        #HPAH1
          "Pyrene",              #HPAH2
          "Benzo(a)anthracene",  #HPAH3
          "Chrysene",            #HPAH4
          "Benzofluoranthene",   #HPAH5
          "Benzo(a)pyrene" ,     #HPAH6
          "Indeno(1,2,3-cd)pyrene", #HPAH7
          "Dibenzo(a,h)anthracene", #HPAH8
          "Benzo(g,h,i)perylene")   #HPAH9

db   <- "L:/DTNA/PGG_SedDB_2018_SQL.mdb"
query_string = "SELECT * FROM ExR_2bSUM_All"
output_suffix = "SUMTOTALPAH_2018_10_15.txt"
con2 <- odbcConnectAccess(db)  ###Make sure dbase is closed####
data <- sqlQuery(con2, query_string) 
data <- data %>% mutate_if(is.factor, as.character)
close(con2)
# filter to LPAH and HPAHs
X <- data %>% 
  filter(txtConstituent %in% pahs,
         !grepl(txtQual, pattern = "T"))
# ADD CLASS
X <- X %>% mutate(CLASS = "TOTAL_PAH")
# RUN SUM
S = general_analytical_summary(X)
# PRODUCE OUTPUT
R <- output_EDD_from_sums_and_ref(x.sum = S, x.ref = X, my_signif = 2)
db_name <- stringr::str_replace(db, "L:/", "") %>%
  stringr::str_replace("/", "_") %>% 
  stringr::str_replace(".mdb", "_")
message( sprintf("Writing to file: %s",paste0(db_name, output_suffix)))
write.table(R, 
            file = paste0(db_name, output_suffix), 
            quote = F, 
            row.names = F, 
            sep= "|")

```


