---
title: "Sum of PAH Groups"
author: "Koshlan Mayer-Blackwell"
date: "January 10, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
require(knitr)
require(tidyr)
require(RODBC)
```
### PAH Summation Script

### Backround

Polycyclic aromatic hydrocarbons (PAHs) are a group of chemicals 
found in petrogenic hydrocarbons and generated from the combustion of 
organic mateiral (e.g. coal, wood, fuels).

Lab data reports contain concentrations of individual chemicals; however, 
regulations are often based on the total aggregate concentration of multiple PAHs. 

Rules and reporting requiremens for summed parameters, particualrly 
when some results are below analytical method detection limits, 
can be agency or site-specific. 

It is also often useful to distinguish total concentrations of 
lower (LPAH) and higher (HPAH) molecular weight PAHs. 

PAHs that are known to be carcinogenic (cPAHs) 
are often summed with a correction factor to report cPAH concentration 
equivalent to the potency of cPAHs—A benzo(a)pyrene (BaP)
Total cPAHs—A benzo(a)pyrene (BaP) equivalent (BaPEq) concentration of a sample 
is calculated by multiplying the carcinogenic PAHs (cPAHs) by their respective 
potency equivalency factors (PEFs), and summing the resulting concentrations. 

### Script Inputs

<li>db - string path to relational databse database</li>
<li>query_string - string SQL query used to select relevant data</li>
<li>output_suffic - string suffix to append to output</li>


```{r inputs}
db   <- "L:/DTNA/PGG_SedDB_2018_SQL.mdb"
query_string = "SELECT * FROM ExR_2bSUM_All"
output_suffix = "SUMPAHs_2018_10_9.txt"  
```

### Script Output Format

PGG Electronic Data Deliverable (EDD)"|" delimited file with LPAH, HPAH, and CPAH PEQ sums

The output of this script is an electronic data deliverable 
with summed results. The EDD produced has the following 
standard fields to comply with database import routines:
  <li> PRJ_NAME       </li>
  <li> EVENT          </li>
  <li> LAB_ID         </li>
  <li> SAMPLE_DT      </li>
  <li> ANALYSIS_METH  </li>
  <li> CONSTITUENT    </li>
  <li> CAS_ID         </li>
  <li> P_CODE         </li>
  <li> SAMPLE_ID      </li>
  <li> EXTRACTED DATE </li>
  <li> ANALYZED_DT    </li>
  <li> DILUTION       </li>
  <li> RESULT         </li>
  <li> UNITS          </li>
  <li> QUAL           </li>
  <li> LIMIT          </li>
  <li> QA             </li>
  <li> COMMENTS       </li>

To accomodate multiple reporting requirements each sum is output according to three definition.
  <li> CPAH_ND_zero - non-detection do not contribute to sum                          </li>
  <li> CPAH_ND_half - non-dections contribute to sum at half detection limit          </li>
  <li> CPAH_ND_full - non-detection contribute to sum at full detection limit         </li>



### Initialize Analyte Definitions as data.frames  

<li> Total PAHs—Total polycyclic aromatic hydrocarbons (PAHs) are the sum of the individual LPAHs and HPAHs. </li>


<li> Total cPAHs—A benzo(a)pyrene (BaP) equivalent (BaPEq) concentration was calculated by multiplying the carcinogenic PAHs (cPAHs) by their respective potency equivalency factors (PEFs), 
and summing the resulting concentrations.  </li>


<li>PAHs classified as carcinogenic are:
benzo(a)anthracene, chrysene, benzo(b)fluoranthene, benzo(k)fluoranthene, benzo(a)pyrene, indeno(1,2,3,-c,d)pyrene, and dibenzo(a,h)anthracene.  </li>


<li> PEFs were assigned according to USEPA (1993) Table 8. Estimated Order of Potential Potencies of Selected PAH Based on Mouse Skin Carcinogenesis "Relative Potency" 2nd column:<li> 


```{r definitions, message = F, warnings = F, echo =T}

cPAH <- data.frame( PAH = c("Benzo(a)anthracene",
                             "Benzo(a)pyrene",
                              "Benzo(b)fluoranthene",
                              "Benzo(k)fluoranthene",
                              "Chrysene",
                              "Dibenzo(a,h)anthracene",
                            "Indeno(1,2,3-cd)pyrene"),
                    PEF = c(0.1, 1, 0.1, 0.01,0.001,1,0.1),
                    stringsAsFactors=FALSE
                    )

PAH <- data.frame( PAH = c("2-Methylnaphthalene", "Acenaphthene", "Acenaphthylene", 
                           "Anthracene", "Fluorene", "Naphthalene", "Phenanthrene",
                           "Fluoranthene", "Pyrene", 
                           "Benzo(a)anthracene", "Chrysene", 
                           "Benzofluoranthene",  "Benzo(a)pyrene" ,
                           "Indeno(1,2,3-cd)pyrene" , "Dibenzo(a,h)anthracene" , 
                           "Benzo(g,h,i)perylene",
                           "Benzo(b)fluoranthene",
                           "Benzo(k)fluoranthene"   ),
                   
                   CLASS = c("LPAH","LPAH","LPAH",
                             "LPAH", "LPAH","LPAH","LPAH",
                             "HPAH", "HPAH",
                             "HPAH", "HPAH",
                             "HPAH", "HPAH",
                             "HPAH", "HPAH",
                             "HPAH", "CPAH", "CPAH"), stringsAsFactors=FALSE )
# PAH LOOKUP will be used below
PAH_LOOKUP <- PAH %>%
  left_join(cPAH) %>% 
  mutate(TYPE = ifelse(!is.na(PEF), "CPAH", "PAH" ))
```
### Define Output Function
```{r functions , message = F, warnings = F, echo = T}
#' output_EDD_from_sums_and_ref
#' 
#' the
#'
#' @param x.sum is a summary data.frame with sums
#' @param x.ref is a reference data.frame with ancillary information about each sampleID
#'
#' @return
#' @export
#'
#' @examples
output_EDD_from_sums_and_ref<- function(x.sum, x.ref){
  
  ref_info <- x.ref %>% group_by(txtSampleID) %>% slice(1) %>% 
    select(-txtConstituent, -txtCASID, -txtQual, -dblLimit, -dblResult)
  
  x1 <- x.sum %>% select(CLASS, txtSampleID, ND_zero, ND_half, ND_full,txtQual, maxLimit) %>%
    gather(summation, dblResult, -CLASS, -txtSampleID, -txtQual, -maxLimit) %>% 
    mutate(txtConstiuent = paste0(CLASS, "_", summation)) %>% ungroup()
  
  x2 <- x1 %>% left_join(ref_info)
  
  x3 <- x2 %>% transmute(PRJ_NAME  = txtProjectName,
                         EVENT     = txtEvent,
                         LAB_ID    = txtLabID,
                         SAMPLE_DT = dtmSampleDate,
                         ANALYSIS_METH  = txtAnalysisMethod,
                         CONSTITUENT    = txtConstiuent,
                         CAS_ID         = NA,
                         P_CODE         = txtPCode,
                         SAMPLE_ID      = txtSampleID,
                         EXTRACTED_DATE = dtmExtractedDate,
                         ANALYZED_DT    = dtmAnalyzedDate,
                         DILUTION       = intDilution,
                         RESULT         = dblResult,
                         UNITS          = txtUnits,
                         QUAL           = txtQual,
                         LIMIT          = maxLimit,
                         QA             = txtQA,
                         COMMENTS       = txtComments)
  
  #change posix date format to MM/DD/YYY hh:mm character
  x3$SAMPLE_DT<-as.character(x3$SAMPLE_DT, format='%m/%d/%y %H:%M')
  x3$EXTRACTED_DATE<-as.character(x3$EXTRACTED_DATE, format='%m/%d/%y %H:%M')
  x3$ANALYZED_DT<-as.character(x3$ANALYZED_DT, format='%m/%d/%y %H:%M')
  #change NAs and NA characters to blanks
  x3[is.na(x3)] <- ""

  return(x3)
}
```

### Database Query and Subsetting

```{r load_data, message = F, warnings = F, echo = T}

# Load data cleanly, subset to PAHs
con2 <- RODBC::odbcConnectAccess(db)  ###Make sure dbase is closed####
data <- sqlQuery(con2, query_string); close(con2)
data <- data %>% dplyr::mutate_if(is.factor, as.character)
d.PAH <- filter(data, txtConstituent %in% PAH_LOOKUP$PAH)
```

### Total LPAH, HPAH
Note that these tidyverse operations work on both analyte results and lab qualifiers. 

```{r lpah_and_hpah_sums, message = F, warnings = F, echo =T}
# CALCULATE LPAH, HPAH SUMS
LPAH.HPAH.sums <- d.PAH %>% select(txtConstituent,txtSampleID,dblResult, dblLimit, txtQual) %>% 
  left_join(PAH_LOOKUP, by = c("txtConstituent" = "PAH"))%>%
  mutate(dblResultzeroND = ifelse(is.na(dblResult), 0*dblLimit, dblResult)) %>%
  mutate(dblResulthalfND = ifelse(is.na(dblResult), 0.5*dblLimit, dblResult)) %>%
  mutate(dblResultfullND = ifelse(is.na(dblResult), 1*dblLimit, dblResult)) %>% 
  filter(CLASS %in% c("LPAH","HPAH")) %>%
  group_by(CLASS, txtSampleID) %>%
  summarise(sumResult_0ND    = sum(dblResultzeroND, na.rm=TRUE), 
                         sumResult_halfND = sum(dblResulthalfND, na.rm=TRUE),
                         sumResult_fullND = sum(dblResultfullND, na.rm=TRUE),
                         maxLimit = max(dblLimit), 
                         n = n(),
                         nd = sum(is.na(dblResult))) %>%
  mutate(`ND_zero`    = ifelse(sumResult_0ND == 0, maxLimit, sumResult_0ND), 
            `ND_half` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_halfND),
            `ND_full` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_fullND))

x.QA <- d.PAH %>%
  left_join(PAH_LOOKUP, by = c("txtConstituent" = "PAH")) %>%
  filter(CLASS %in% c("LPAH","HPAH")) %>% 
  group_by(txtSampleID, CLASS) %>%
  summarise( txtQualT = toString( unique(txtQual), sep = "")) 
# [x.flags] collects all unique flags for a given sample (e.g BJq)
x.flags <- stringr::str_replace_all(x.QA$txtQualT, pattern =",", replacement = "") %>%
  #stringr::str_replace_all( pattern ="C", replacement = "") %>% 
  stringr::str_replace_all( pattern =" +", replacement = "-") %>% 
  stringr::str_split(pattern = "-") %>%
  lapply(function(x) unique(x)) %>%
  lapply(function(x) sort(x)) %>%
  lapply(function(x) toString(x)) %>% 
  stringr::str_replace_all( pattern =",", replacement = "") %>%
  stringr::str_replace_all( pattern =" ", replacement = "") %>%
  stringr::str_replace_all( pattern ="U", replacement = "") %>%
  stringr::str_replace_all( pattern ="C", replacement = "") %>%
  paste0("T") %>%
  stringr::str_replace_all( pattern ="NA", replacement = "")

if (length(x.flags) == dim(LPAH.HPAH.sums)[1]){
  LPAH.HPAH.sums$txtQual <- x.flags
}else{
    stop("Error due to mismatch in number of sums and total flags")
}
```

### Total cPAHs
Total cPAHs as benzo(a)pyrene (BaP) equivalent (BaPEq) concentration was calculated by multiplying the carcinogenic PAHs (cPAHs) by their respective potency equivalency factors (PEFs)
```{r cpah_sums, message = F, warnings = F, echo =T}
CPAH.sums <- d.PAH %>% select(txtConstituent,txtSampleID,dblResult, dblLimit, txtQual) %>% 
  left_join(PAH_LOOKUP, by = c("txtConstituent" = "PAH"))%>%
  mutate(dblResultzeroND = ifelse(is.na(dblResult), 0*dblLimit, dblResult*PEF)) %>%
  mutate(dblResulthalfND = ifelse(is.na(dblResult), 0.5*dblLimit, dblResult*PEF)) %>%
  mutate(dblResultfullND = ifelse(is.na(dblResult), 1*dblLimit, dblResult*PEF)) %>% 
  filter(TYPE %in% c("CPAH")) %>%
  group_by(TYPE, txtSampleID) %>%
  summarise(sumResult_0ND    = sum(dblResultzeroND, na.rm=TRUE), 
            sumResult_halfND = sum(dblResulthalfND, na.rm=TRUE),
            sumResult_fullND = sum(dblResultfullND, na.rm=TRUE),
            maxLimit = max(dblLimit), 
            n = n(),
            nd = sum(is.na(dblResult))) %>%
  mutate(`ND_zero`    = ifelse(sumResult_0ND == 0, maxLimit, sumResult_0ND), 
         `ND_half` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_halfND),
         `ND_full` = ifelse(sumResult_0ND == 0, maxLimit, sumResult_fullND)) %>%
  mutate(CLASS = TYPE) %>% ungroup()


x.QA <- d.PAH %>%
  left_join(PAH_LOOKUP, by = c("txtConstituent" = "PAH")) %>%
  filter(TYPE %in% c("CPAH"))  %>% 
  group_by(TYPE, txtSampleID) %>%
  summarise( txtQualT = toString( unique(txtQual), sep = "")) 
# [x.flags] collects all unique flags for a given sample (e.g BJq)
x.flags <- stringr::str_replace_all(x.QA$txtQualT, pattern =",", replacement = "") %>%
  #stringr::str_replace_all( pattern ="C", replacement = "") %>% 
  stringr::str_replace_all( pattern =" +", replacement = "-") %>% 
  stringr::str_split(pattern = "-") %>%
  lapply(function(x) unique(x)) %>%
  lapply(function(x) sort(x)) %>%
  lapply(function(x) toString(x)) %>% 
  stringr::str_replace_all( pattern =",", replacement = "") %>%
  stringr::str_replace_all( pattern =" ", replacement = "") %>%
  stringr::str_replace_all( pattern ="U", replacement = "") %>%
  stringr::str_replace_all( pattern ="C", replacement = "") %>%
  paste0("T") %>%
  stringr::str_replace_all( pattern ="NA", replacement = "")

if (length(x.flags) == dim(CPAH.sums)[1]){
  CPAH.sums$txtQual <- x.flags
}else{
  stop("Error due to mismatch in number of sums and total flags")
}
```



### Execute the outputs function to write external EDD file #
```{r output_edd, message = F, warnings = F, echo = T}
x4 <- output_EDD_from_sums_and_ref(x.sum = LPAH.HPAH.sums,
                               x.ref = d.PAH)
x5 <- output_EDD_from_sums_and_ref(x.sum = CPAH.sums,
                                   x.ref = d.PAH)
x5$COMMENTS <- "CPAH (BaPEq) concentration was calculated by multiplying the carcinogenic PAHs (cPAHs) by their respective potency equivalency factors (PEFs)"

x6 <- rbind(x4,x5)
x6$RESULT <- signif(x6$RESULT,2 )
db_name <- stringr::str_replace(db, "L:/", "") %>% 
  stringr::str_replace("/", "_") %>% 
  stringr::str_replace(".mdb", "_")

write.table(x6, 
            file = paste0(db_name, output_suffix), 
            quote = F, 
            row.names = F, 
            sep= "|")
```

### Preview Output
```{r preview, message = F, warnings = F, echo = T}
head(x6)
```


