---
title: "Sum of PCB Congeners"
author: "Koshlan Mayer-Blackwell"
date: "January 11, 2019"
output: html_document
---
### Setup R environment
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
require(knitr)
require(tidyr)
require(RODBC)
```

### INPUTS: Script Specific Database Path, Query, and Suffix

These inputs are specific to the database.

This script assumes that the will be recovered from and existing table (e.g. ExR_2bSUM_All)

```{r inputs}
# Load data cleanly Load
db <- "L:/DTNA/PGG_SedDB_2018_SQL.mdb"
query_string = "SELECT * FROM ExR_2bSUM_All"
output_suffix = "PCBcTotal.txt"
```

### FUNCTIONS 
<li>output_EDD_from_sums_and_ref </li>
<li>convert_to_coelution_reporting_format </li>
<li>proccess </li>

```{r functions, echo=TRUE, message = FALSE}
#' output_EDD_from_sums_and_ref
#'
#' @param x.sum a data.frame with column names
#'   "txtSampleID", 
#'   "sumResult_0ND", 
#'   "sumResult_halfND", 
#'   "sumResult_fullND", 
#'   "maxLimit", 
#'   "n", 
#'   "nd", 
#'   "ND_zero", 
#'   "ND_half", 
#'   "ND_full", 
#'   "txtQual", 
#'   "CLASS"
#' @param x.ref a data.frame with other sample information which will be joined. It must contain the following headers to avoid an error
#'    txtProjectName
#'    txtEvent
#'    txtLabID
#'    dtmSampleDate
#'    txtAnalysisMethod
#'    txtConstiuent
#'    txtPCode
#'    txtSampleID
#'    dtmExtractedDate
#'    dtmAnalyzedDate
#'    intDilution
#'    dblResult
#'    txtUnits
#'    txtQual
#'    maxLimit
#'    txtQA
#'    txtComments
#'
#' @return x3 - data.frame with summation results in EDD format
output_EDD_from_sums_and_ref<- function(x.sum, x.ref){
  
  ref_info <- x.ref %>% group_by(txtSampleID) %>% slice(1) %>% 
    select(-txtConstituent, -txtCASID, -txtQual, -dblLimit, -dblResult)
  
  x1 <- x.sum %>% select(CLASS, txtSampleID, ND_zero, ND_half, ND_full,txtQual, maxLimit) %>%
    gather(summation, dblResult, -CLASS, -txtSampleID, -txtQual, -maxLimit) %>% 
    mutate(txtConstiuent = paste0(CLASS, "_", summation)) %>% ungroup()
  
  x2 <- x1 %>% left_join(ref_info)
  
  x3 <- x2 %>% transmute(PRJ_NAME  = txtProjectName,
                         EVENT     = txtEvent,
                         LAB_ID    = txtLabID,
                         SAMPLE_DT = dtmSampleDate,
                         ANALYSIS_METH  = txtAnalysisMethod,
                         CONSTITUENT    = txtConstiuent,
                         CAS_ID         = NA,
                         P_CODE         = txtPCode,
                         SAMPLE_ID      = txtSampleID,
                         EXTRACTED_DATE = dtmExtractedDate,
                         ANALYZED_DT    = dtmAnalyzedDate,
                         DILUTION       = intDilution,
                         RESULT         = dblResult,
                         UNITS          = txtUnits,
                         QUAL           = txtQual,
                         LIMIT          = maxLimit,
                         QA             = txtQA,
                         COMMENTS       = txtComments)
  
  #change posix date format to MM/DD/YYY hh:mm character
  x3$SAMPLE_DT<-as.character(x3$SAMPLE_DT, format='%m/%d/%y %H:%M')
  x3$EXTRACTED_DATE<-as.character(x3$EXTRACTED_DATE, format='%m/%d/%y %H:%M')
  x3$ANALYZED_DT<-as.character(x3$ANALYZED_DT, format='%m/%d/%y %H:%M')
  #change NAs and NA characters to blanks
  x3[is.na(x3)] <- ""
  return(x3)
}

#' convert_to_coelution_reporting_format
#' 
#' PCB-100, with flag C93 implies that PCB-100 coelutes with C93, we want a single concentration for  PCB-93+100 to avoid double counting
#'    
#'    This function converts the Test America Flag format to the classic pattern PCB-10+14#'    input must contain:
#'    txtConstituent
#'    txtQual with Coelution format as "C([0-9]+)"
#' 
#' @param df 
#'
#' @return output (df with a new column PCBs that includes the coeluting congeners with plus signs, like PCB-4+10 )
convert_to_coelution_reporting_format <- function(df){
  m <- as.data.frame(cbind(df$txtConstituent, stringr::str_match(df$txtQual, pattern = "C([0-9]+)")), stringsAsFactors = F)
  m$V3 <- as.numeric(m$V3)
  m$V4 <- as.numeric(stringr::str_replace(m$V1, pattern = "PCB-", replacement = ""))
  m$ORIGINAL <- df$txtConstituent
  m <- m[order(m$V4),]
  for(i in 1:dim(m)[1]){
    coelution <- m$V3[i]
    if (!is.na(coelution)){
      ind = which(m$V4 == coelution)
      m$V1[ind] <- paste(m$V1[ind], m$V4[i], sep = "+")
    }
  }
  table <- data.frame(ORIGINAL = m$ORIGINAL, PCB= m$V1 , stringsAsFactors = F)
  output <-  left_join(df, table, by = c("txtConstituent" = "ORIGINAL"))
  return(output)
}

#' process
#' this is a wrapper function for convert_to_coelution_reporting_format(), 
#  it also sorts congener in ascending order
#
#' @param df 
#'
#' @return 
process <- function(df){
  df2 <- convert_to_coelution_reporting_format(df) # Produces new PCB label PCB-Z+X+Y
  df3 <- df2[!grepl(df2$txtQual, pattern = "C[0-9]+"),] # Drops all samples that were reported as a coelution with Z
  df3$SORTBY <- as.numeric(stringr::str_replace(df3$txtConstituent, pattern = "PCB-", replacement = "")) # extracts a numeric congener values to sort by
  df4 <- arrange(df3, SORTBY) # sorts so PCB-20 comes before PCB-100 
  return(df4)
}


```

### Query Database
```{r query}
con2 <- odbcConnectAccess(db)  ###Make sure dbase is closed####
data<-sqlQuery(con2, query_string)
close(con2)
```
### Subset the data to PCB congeners only
```{r subset_to_pcbs}
# [PCB] Select only PCB-Congener Results
PCB <- data %>% mutate_if(is.factor, as.character) %>% 
  filter(txtAnalysisMethod == "E1668A", grepl("PCB-*", txtConstituent))
# process each sample individually by breaking the data.frame into a series of data.frames by each SampleID
l <- split(PCB, f =  PCB$txtSampleID)
l2 <- lapply(l, function(i) process(i))
dPCB <- do.call(rbind, l2)
```

### Generate the x.sum object
```{r xsums}
# [x.sum] aggregate according to RI summary rules
x.sum <- dPCB %>%
  group_by(txtSampleID) %>%
  # add a column which is dblResult or half dblLimit, if ND
  mutate(dblResultzeroND = ifelse(is.na(dblResult), 0*dblLimit, dblResult)) %>%
  mutate(dblResulthalfND = ifelse(is.na(dblResult), 0.5*dblLimit, dblResult)) %>%
  mutate(dblResultfullND = ifelse(is.na(dblResult), 1*dblLimit, dblResult)) %>% 
  # sum over all the congeners
  summarise(sumResult_0ND    = sum(dblResultzeroND, na.rm=TRUE), 
            sumResult_halfND = sum(dblResulthalfND, na.rm=TRUE),
            sumResult_fullND = sum(dblResultfullND, na.rm=TRUE),
            maxLimit = max(dblLimit, na.rm = T), 
            n = n(), 
            nd = sum(dblResult <= dblLimit)) %>%
  # If all results are non-detect use highest single dection limit
  mutate(`ND_zero`    = ifelse(sumResult_0ND == 0, maxLimit, sumResult_0ND), 
          `ND_half` = ifelse(sumResult_halfND == 0, maxLimit, sumResult_halfND),
          `ND_full` = ifelse(sumResult_fullND == 0, maxLimit, sumResult_fullND))
```
### Generate x.flags objects
```{r xflags}
# [x.QA] outputs all unique flag combinations for a given sample (e.g. J q, J, J B, J C q)
x.QA <- dPCB %>%
  group_by(txtSampleID) %>%
  summarise( txtQualT = toString( unique(txtQual), sep = "")) 

# [x.flags] collects all unique flags for a given sample (e.g BJq)
x.flags <- stringr::str_replace_all(x.QA$txtQualT, pattern =",", replacement = "") %>% # removes the comma separating flags
  # replace blank spaces with dashes
  stringr::str_replace_all( pattern =" +", replacement = "-") %>% 
  # split individual flags into list
  stringr::str_split(pattern = "-") %>%
  # only retain unique flags
  lapply(function(x) unique(x)) %>%
  # sort flags
  lapply(function(x) sort(x)) %>%
  # combine all flags into one long string
  lapply(function(x) toString(x)) %>% 
  # remove comas
  stringr::str_replace_all( pattern =",", replacement = "") %>%
  # remove spaces
  stringr::str_replace_all( pattern =" ", replacement = "") %>%
  # removed U flag
  stringr::str_replace_all( pattern ="U", replacement = "") %>%
  # remove C flag
  stringr::str_replace_all( pattern ="C", replacement = "") %>%
  # add one Total T flag
  paste0("T") %>%
  # remove any NA strings 
  stringr::str_replace_all( pattern ="NA", replacement = "") %>%
  # split the string into individual characters
  strsplit("") %>% 
  # keep only unique letters
  lapply(function(x) unique(x)) %>%
  lapply(function(x) paste(x, collapse = '')) %>%
  unlist()
```


### Combine Sums and Summary Flags
```{r combine_sums_and_flags}
# check to make sure flags and summary data have teh same row dimensions
if (length(x.flags) != dim(x.sum)[1]){
  stop("Data mismatch between summary data and data flags list")
}else{
  x.sum$txtQual <- x.flags
  #print("Data summary and flags dimensions match. Flags appended.")
}
```

### Generate Output
```{r format_output}
x.sum$CLASS = "Total_PCB"
results <- output_EDD_from_sums_and_ref(x.sum=x.sum, x.ref=dPCB )
results$RESULT <- signif(results$RESULT, 2)
db_name <- stringr::str_replace(db, "L:/", "") %>%
  stringr::str_replace("/", "_") %>% 
  stringr::str_replace(".mdb", "_")
```

### Preview Output
```{r preview_output}
head(results)
```

### Write output to txt file
```{r write output to file}
write.table(results, 
            file = paste0(db_name, output_suffix), 
            quote = F, 
            row.names = F, 
            sep= "|")
```
